---
title: "Worked Example: Female 5xFAD Time Course with Centroid-Based Trajectory Collapse"
author: "Saren M. Springer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked Example: Female 5xFAD Time Course}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

# Overview

This vignette documents a complete WATER analysis of female wildtype samples from a 5xFAD aging dataset. The example contains four ages—4, 8, 12, and 18 months—with three biological replicates per age. WATER was run using its default centroid-based method for collapsing related clusters into trajectories.

The purpose of this vignette is to show:

1. how to organize the input data;
2. how to launch a complete WATER run;
3. which interactive decisions were made during clustering and refinement;
4. how to interpret the audit and stability outputs; and
5. where the final results are saved.

This is a workflow example rather than a biological interpretation of the dataset.

# Dataset structure

The expression matrix contained 55,351 input features and 12 samples. Samples were ordered by age, with three replicates at each timepoint.

| WATER timepoint | Biological age | Replicates | Sample names |
|---|---:|---:|---|
| T1 | 4 months | 3 | WT1_F_4M, WT2_F_4M, WT3_F_4M |
| T2 | 8 months | 3 | WT1_F_8M, WT2_F_8M, WT3_F_8M |
| T3 | 12 months | 3 | WT1_F_12M, WT2_F_12M, WT3_F_12M |
| T4 | 18 months | 3 | WT1_F_18M, WT2_F_18M, WT3_F_18M |

Two input tables were used:

- an expression matrix, with feature identifiers in the first column and samples in the remaining columns;
- a log2 fold-change matrix aligned to the same feature identifiers.

A feature list was also supplied through `feature_list = "filter.txt"`.

# Load WATER and the input files

During package development, the WATER runner can be sourced directly.

```{r load-inputs}
data <- read.csv("WT_female_Alz_exp_data.csv")
log2fc <- read.csv("WT_female_Alz_log2fc_data.csv")

source("Running_WATER.R")
```


# Run WATER

The complete analysis was launched with `run_water()`.

```{r run-water}
water <- run_water(
  data = data,
  replicates = list(
    T1 = 3,
    T2 = 3,
    T3 = 3,
    T4 = 3
  ),
  log2fc = log2fc,
  preprocess_params = list(
    modality = "RNA",
    output_dir = "WT_female_Alz",
    save_outputs = TRUE,
    tpm_threshold = 1.0,
    feature_list = "filter.txt"
  )
)
```

The output directory used in this example was:

```text
WT_female_Alz/
```

WATER created stage-specific subdirectories containing tables, R objects, and plots.

# Preprocessing results

WATER first initialized the sample structure, filtered features, aligned the expression and log2 fold-change inputs, normalized the expression values, and performed PCA.

| Preprocessing step | Features retained |
|---|---:|
| Raw input | 55,351 |
| TPM threshold, timepoint mean ≥ 1 | 18,246 |
| Feature-list filter | 13,147 |
| Differentially abundant features | 1,980 |
| Non-differentially abundant features | 11,167 |

The 1,980 differentially abundant features were passed to temporal clustering.

# Initial clustering and trajectory labeling

WATER evaluated candidate cluster numbers from `k = 2` through `k = 30` using silhouette analysis and generated heatmaps for candidate solutions.

The recommended values were:

```text
k = 3, 8, 16, 17
```

For this analysis, `k = 8` was selected based on inspection of the resulting hierarchical clustering.

| Initial cluster | Features |
|---:|---:|
| 1 | 1,196 |
| 2 | 123 |
| 3 | 171 |
| 4 | 75 |
| 5 | 114 |
| 6 | 115 |
| 7 | 117 |
| 8 | 69 |

## Centroid-based trajectory collapse

At the trajectory-labeling prompt, the default centroid method was selected:

```text
Centroid clustering, r_threshold = 0.8
```

WATER compared the mean z-scored expression profile of each cluster and grouped clusters whose centroid profiles met the correlation threshold. At *r* = 0.8, the eight initial clusters remained as eight distinct trajectories in this dataset.

WATER then wrote a rename template:

```text
label_classic/tables/trajectory_rename_template.csv
```

The `Biological_label` column was completed manually, and the edited file was supplied back to WATER. The initial eight trajectories were:

| Trajectory | Features |
|---|---:|
| 4M_High_and_18M_High | 1,196 |
| 18M_High | 171 |
| 12M_High | 123 |
| 4M_High | 117 |
| Down-Up-Down | 115 |
| 18M_Low | 114 |
| Late_High | 75 |
| 8M_High | 69 |

# Initial audit

WATER next evaluated assignment stability using leave-one-out analysis and a correlation-based audit.

## Initial leave-one-out analysis

Each of the 12 samples was removed once, and the clustering solution was reevaluated.

| Metric | Initial result |
|---|---:|
| Mean adjusted Rand index | 0.289 |
| Cluster retention | 50.5% |
| Trajectory retention | 50.5% |

## Initial correlation audit

Each feature was compared with the centroid of its assigned trajectory and with competing trajectories.

| Audit decision | Features | Percent |
|---|---:|---:|
| Keep original | 1,533 | 77.4% |
| Reassigned | 210 | 10.6% |
| Weak | 88 | 4.4% |
| Ambiguous | 149 | 7.5% |

The weak and ambiguous feature sets were carried into dedicated refinement steps rather than being discarded.

# Refinement of ambiguous features

The 149 ambiguous features were reclustered independently. WATER recommended:

```text
k = 2, 4, 7
```

For this run, `k = 4` was selected. The default centroid-based collapse method was again used with `r_threshold = 0.8`. All four ambiguous clusters remained distinct before biological renaming.

| Ambiguous trajectory | Features |
|---|---:|
| 18M_High | 92 |
| 4M_High_and_18M_High | 28 |
| 12M_High | 16 |
| 18M_Low | 13 |

The ambiguous rename template was saved to:

```text
recluster_ambiguous/tables/ambiguous_rename_template.csv
```

# Refinement of weak features

The 88 weak features were also reclustered independently. WATER recommended:

```text
k = 2, 8, 10
```

For this run, `k = 8` was selected. The default centroid-based collapse method was used with `r_threshold = 0.8`. The eight weak clusters remained distinct before renaming.

| Weak trajectory | Features |
|---|---:|
| 8M_Low | 28 |
| 12M_Low | 17 |
| 4M_Low | 16 |
| Late_High | 11 |
| Early_High | 8 |
| 4M_High | 6 |
| Middle_High | 2 |

The weak rename template was saved to:

```text
recluster_weak/tables/weak_rename_template.csv
```

# Final integrated trajectories

After integration of original, reassigned, ambiguous, and weak features, all 1,980 differentially abundant features were represented in the final trajectory set.

| Final trajectory | Features |
|---|---:|
| 4M_High_and_18M_High | 904 |
| 18M_High | 296 |
| 4M_High | 196 |
| Down-Up-Down | 143 |
| 12M_High | 136 |
| 18M_Low | 119 |
| Late_High | 66 |
| 8M_High | 49 |
| 8M_Low | 28 |
| 12M_Low | 17 |
| 4M_Low | 16 |
| Early_High | 8 |
| Middle_High | 2 |
| **Total** | **1,980** |

# Secondary correlation validation

WATER repeated the correlation audit after refinement.

| Final audit decision | Features | Percent |
|---|---:|---:|
| Stable | 1,908 | 96.4% |
| Reassign candidate | 49 | 2.5% |
| Low confidence | 23 | 1.2% |

The per-trajectory correlation summary was:

| Final trajectory | Mean correlation | Median correlation | Features |
|---|---:|---:|---:|
| 4M_High_and_18M_High | 0.941 | 0.954 | 890 |
| 18M_High | 0.956 | 0.971 | 285 |
| 4M_High | 0.942 | 0.954 | 185 |
| Down-Up-Down | 0.919 | 0.935 | 139 |
| 12M_High | 0.940 | 0.956 | 131 |
| 18M_Low | 0.917 | 0.929 | 122 |
| Late_High | 0.883 | 0.891 | 60 |
| 8M_High | 0.933 | 0.951 | 49 |
| 8M_Low | 0.916 | 0.969 | 38 |
| 4M_Low | 0.836 | 0.881 | 25 |
| 12M_Low | 0.910 | 0.927 | 15 |
| Early_High | 0.888 | 0.930 | 13 |
| Middle_High | 0.777 | 0.795 | 5 |

# Secondary leave-one-out validation

The refined trajectory assignments were reevaluated across all 12 leave-one-out iterations.

| Metric | Secondary result |
|---|---:|
| Mean trajectory retention | 78.6% |
| Mean reassignment | 21.4% |
| Mean low-confidence classification | 1.5% |

The increase in retention from the initial to the secondary analysis reflects the effect of WATER's reassignment and targeted reclustering steps.

# Output directory guide

The main output folders generated during this analysis were:

```text
WT_female_Alz/
|-- initialize/
|-- filter_expression/
|-- filter/
|-- normalize/
|-- PCA/
|-- cut_classic/
|-- label_classic/
|-- loo_initial/
|-- correlate_initial/
|-- recluster_ambiguous/
|-- recluster_weak/
|-- correlate_secondary/
|-- loo_secondary/
`-- final/
```

Each analysis stage contains one or more of the following:

```text
tables/    Tabular outputs in .xlsx, .csv, or .txt format
objects/   Serialized R objects in .rds format
plots/     Diagnostic or final visualizations
```

# Session information

Record the package and R versions used for the analysis:

```{r session-info}
sessionInfo()
```
